%
% Formative essay 1
%
% Reuben Thomas
%


\documentclass[english]{scrartcl}
\usepackage{babel,newpxtext,url,booktabs}
\usepackage[utf8x]{inputenc}



% Alter some default parameters for general typesetting

\frenchspacing
%TC:macro \subtitle [header]


% New commands



\begin{document}

\title{How does induction differ from deduction?\\Why is induction important to reasoning\\in science?}
\subtitle{Formative Essay 1\\\input{Formative essay 1.wc}words}
\author{Reuben Thomas}
\date{23rd October 2021}
\maketitle

Inductive inference is generally held to be an entirely distinct mode of reasoning from mathematical induction. In this essay I shall show that, on the contrary, they are structurally similar, and that this similarity illuminates the importance of inductive inference to science.

Induction and deduction are methods of argument that use inference to reason from premisses to conclusion. Deduction applies the rules of some logic to the premisses to reach the conclusion: the reasoning is tautological. In an inductive inference, the conclusion is merely supported by the premisses. Deduction is universally accepted as a valid mode of reasoning.

Induction’s importance to science is that it permits generalization: without it, we would have only knowledge by acquaintance and propositional knowledge. (An omniscient being could work out general laws, because their laws would not need to predict anything, as they would already be in possession of all possible observations.) Induction has not been formulated in a logically rigorous way; the classic statement of “the problem of induction” is given by Hume~\cite{hume1978}. For example, I may observe many white swans, and conclude that all swans are white; but Hume says “the sheer number of impressions has in this case no more effect than if we confined ourselves to one only”: that is, concluding that all swans are white from seeing one white swan is just as valid as reaching the conclusion from seeing a hundred. (Popper might remind us that it only takes a single black swan to falsify the proposition.)

Inductive reasoning and mathematical induction are usually treated as unrelated modes of reasoning that unfortunately share a name: for example, Wikipedia’s article “Mathematical induction”~\cite{wiki:mathematical-induction} says “Although its name may suggest otherwise, mathematical induction should not be confused with inductive reasoning as used in philosophy.” Mathematical induction is indeed a type of deductive reasoning, but, as I will explain, it is nonetheless a form of induction; the difference between mathematical induction and inductive reasoning is more subtle than it first appears, and illuminates the difference between mathematics and science.

Mathematical induction is a proof technique used to prove that some proposition $P(n)$ holds for every natural number $0, 1,…$. Other forms of mathematical induction apply to more general “well-founded” structures. A “proof by induction” consists of two cases: the \textbf{base case}, which proves $P(0)$, and the \textbf{induction step}, which proves that $P(n)$ implies $P(n+1)$. Taken together with the axiom of induction, these prove $P(n)$ for all $n$.

A classic example is the proof that the sum of the first $n$ natural numbers is $\frac{n(n+1)}{2}$. So, we must prove, for all $n$, $P(n): 0+1+…n = \frac{n(n+1)}{2}$.

First, the base case: $P(0)$ is the proposition $0 = \frac{0(0+1)}{2}$. We can simplify the right-hand side to $\frac{0×1}{2}$, which is $\frac{0}{2}$, which is $0$. So, we have proved the base case.

The induction step is to prove that $P(n)\supset P(n+1)$. In other words, we can assume $P(n)$, and we must prove $P(n+1)$. Substituting $n+1$ for $n$ in the definition of $P(n)$, we see that $P(n+1)$ is the proposition $0+1+…n+(n+1) = \frac{(n+1)(n+2)}{2}$. We can use our assumption of $P(n)$ to substitute $\frac{n(n+1)}{2}$ into the left-hand side, obtaining: $\frac{n(n+1)}{2}+(n+1) = \frac{(n+1)(n+2)}{2}$. Now, we multiply out the right-hand bracket of the right-hand side: $\frac{n(n+1)}{2}+(n+1) = \frac{(n+1)n}{2} + \frac{(n+1)2}{2}$. By rearranging the left-hand fraction and cancelling the $2$s in the right-hand fraction, we obtain: $\frac{n(n+1)}{2}+(n+1) = \frac{n(n+1)}{2} + (n+1)$. We can see that the left-hand side and right-hand side are now identical, so we have proved $P(n+1)$. Hence, “by induction”, we have proved $P(n)$ for all $n$.

To see how this works for all $n$, go back to $P(0)$, which we proved above. In the induction step, we showed that if $P(0)$ is true, then $P(1)$ is true. We can now repeat the induction step: assuming $P(1)$ is true, $P(2)$ must be true; and we can continue to prove $P(n)$ for any $n$, because for any desired $n$ we will reach it eventually by counting up from $0$.

The proofs of the base case and induction step are indeed valid deductive arguments. However, to complete the proof we must accept the axiom of induction: just as in the case of inductive inference, the validity of our proof by (mathematical) induction rests on our accepting the validity of induction.

The following table shows the parallel structure of inductive inference and mathematical induction:

\begin{center}
  \begin{tabular}{cc}
    \textbf{Inductive inference} & \textbf{Mathematical induction}\\
    Base case & Observations\\
    Induction step & Description of causal relation\\
    Axiom of induction & “Axiom of induction”\\
  \end{tabular}
\end{center}

This puts the question, “Why is induction important to reasoning in science?” in a new light: the answer is, because it lets us generalize, \emph{just as mathematical induction does}. Far from being a poor relative of mathematical induction, unjustified as a valid mode of reasoning, merely, in Reichenbach’s words, “the best wager we can lay”~\cite{reichenbach1938eap}, inductive inference is the analogue in the realm of physics of the axiom of induction. Just as mathematical induction allows us to make proofs about all natural numbers from observations about a few, inductive inference is the essential tool we grasp to turn our individual observations into general laws.

\bibliographystyle{plain}
\bibliography{philosophy}

                                                                                                                                                                                        %TC:ignore                                                                                                                                \section*{Acknowledgement}

I thank Alistair Turnbull for critically reading this essay, and in particular for restoring to my attention the axiom of induction, which I had forgotten, except for a vague impression that mathematical induction was not as watertight as the consensus in the seminar held it to be!
%TC:endignore

\end{document}

% LocalWords:  Reichenbach
